----
author: "**Garima Garg**"
header-includes:
   - \usepackage{bbm}
   output: html_document
    fontsize: 8pt  
----
****************************************************************************************************
# Least Squares Solutions for Sparse Matrices

Here, the matrix A is sparse, and we are trying to find X that minimizes $\left | AX-b \right |_{2}^{2}$,   $A \epsilon R^{m*n}$ where $m\geq n$

Following methods were used to find the least squares solution when A is sparse and m>n. <br/>
- **Singular Value Decomposition** <br/>
- **Successive Over Relaxation (SOR)** <br/>
- **Ridge Regression (Constrained least squares)**


****************************************************************************************************
#Singular Value Decomposition <br/>

$$A=U \Sigma V^T $$  <br/>
$$x^{*} = V \Sigma^{-1} U^T b $$

```{r}
#simulate data
m <- 10000
n <- 500

A <- matrix(rnorm(m * n), m, n)
iz <- sample(1:(m * n),
             size = m * n * 0.85,
             replace = FALSE)
A[iz] <- 0

beta <- rnorm(n)
b <- A %*% beta + rnorm(m)

#rank of matrix
rank <- qr(A)$rank

# Start the clock!
ptm <- proc.time()

svd.data <- svd(A)
result <- svd.data$v %*% diag(1/svd.data$d) %*% t(svd.data$u)
x <- round(result %*% b, digits=4)
```
**Shows system time for SVD computation** 

```{r}

proc.time() - ptm

predict <- A %*% x 
residuals <- b- A %*% x
```
**Error using SVD method** 
```{r, fig.width=4, fig.height=4}
(error <- mean(residuals^2))
plot(b ~ predict, main="Predicted vs actual values using SVD")

```

********************************************************************************************************************************
#Successive Over Relaxation (SOR) <br/>

SOR is an iterative method, and is a variant of the Gauss–Seidel method for solving a linear system of equations, resulting in faster convergence. Since A is not a square matrix, $A^{T}A$ is used instead.
$$ x^{(k+1)} = (1-\omega) x^{k}+\omega(D+L)^{-1}(A^{T} b-U x^{k})$$
$$ A^{T}A = L+U+D $$

```{r}
#function to implement SOR method. There wasn't any package in R that implements SOR.
SOR <- function(B, b, x0, iter, omega){
  
  B <- t(A) %*% A 
  D <- diag(diag(B))
  L <- B
  L[upper.tri(L, diag = TRUE)] <-0
  B[lower.tri(B, diag = TRUE)] <-0
  U <- B
  
  inv.dl <- solve(D+L)
  #x <- rep(0, iter)
  #x[1] <- x0
  x <- x0
  
  for (i in 1:(iter-1))
    x <- (1-omega)*x + omega* (inv.dl %*% (t(A) %*% b - U %*% x))
  
  return (x)
}

#sequence of omega's
omega <- seq(-1, 2.5, by=0.2)
error <- rep(0, length(omega))

# Start the clock!
ptm <- proc.time()

#find the best omega that minimizes error
for(i in 1:length(omega))
  {
  x <- SOR(A, b, x0= rep(0.2,500), iter=50, omega[i])
  residuals <- b- A %*% x
  error[i] <- round(mean(residuals^2), digits=3)
  }
```
**System time to run SOR method**
```{r, fig.width=4, fig.height=4}
proc.time() - ptm

plot(error~omega, main="find best omega that minimize error for SOR method", col = "dark red")
omega.position <- which.min(error)
```
**Error using SOR method** 
```{r, fig.width=4, fig.height=4}
(error <- min(error))

#using the best omega to get the predicted value and plotting the results
final.x <- SOR(A, b, x0= rep(0.2,500), iter=50, omega[omega.position])
predict <- A %*% final.x 
plot(b ~ predict, main="Predicted vs actual values using SOR method")
```

********************************************************************************************************************************************
# Ridge Regression <br/>

This is constrained form of least squares and widely used in statistics. In this, we find X that minimizes  $\left | AX-b \right |_{2}^{2}$ subject to additional constraint $\sum_{i=1}^{j}X_{j}^2 < c$ where c is some constant c>0. <br/>
This is equivalent to minimizing $\sum_{i=1}^{m} (AX-b)^{2}$ + $\lambda \sum_{j=1}^{n}X_{j}^2$, so in this method we find the optimal lambda that reduces the error.  <br/> <br/>

This is solved by using R package glmnet. Additionally, glmnet handles sparse data, and hence more efficient.

```{r, fig.width=4, fig.height=4}
library(glmnet)
A <- Matrix(A, sparse=TRUE)

# Start the clock!
ptm <- proc.time()

fit = glmnet(x=A, y=b, alpha=0) # alpha=0 suggests ridge regression, there are other forms of constrained least squares too
cvfit = cv.glmnet(x=A,y=b)
plot(cvfit, main="lambda that minimizes error for Ridge regression")

```
<br/>
**System time to run ridge regression**
```{r}
proc.time() - ptm

predict <- predict(cvfit, newx = A, s = "lambda.min")
residuals <- b - predict
```
**Error using Ridge Regression method** 
```{r, fig.width=4, fig.height=4}
(error <- mean(residuals^2))
plot(b ~ predict, main="Predicted vs actual values using Ridge Regression")

```

************************************************************************************************************************************
#Conclusions <br/>
SVD and Ridge regression(constrained least squares) performs better than Successive Over Relaxation in terms of system time. Error is almost same for all the three methods.  <br/>

#Future Work 
Advanced iterative methods and other forms of constrained least squares(like L1 penalty) can be used when matrix A is sparse and m>n. 

